\large{\bf Abnormal quantitative genetics? }

### 2023-07-04 writeup.tex
paper plots
.

Now let us choose the distribution of allelic effects so that
the trait distributions converge to a nontrivial limit after rescaling by a different exponent $\alpha$.
Can the property of independence be preserved?
The following calculations suggest the answer is ``no``.

\begin{lem} \label{lem:nongaussian_nonindep}
    \comment{SOMETHING LIKE THE FOLLOWING?}
    Suppose that the allelic effects are such that
    $$
        \lim_{M \to \infty} \log \E[ \prod_{\ell=1}^M e^{it\eta^{[i]}} ]
        = c |t|^\alpha .
    $$
    Then $(Z_M, Z_M^{[1]}, Z_M^{[2]})$ converges in distribution
    to $(Z, Z^{[1]}, Z^{[2]})$ for which each of the three traits
    marginally has a standard $\alpha$-stable distribution,
    but the conditional distribution of $R = Z - (Z^{[1]} - Z^{[2]}))/2$
    given $Z^{[1]}=z_1$ and/or $Z^{[2]}=z_2$ depends nontrivially on $z_1$ and $z_2$.
\end{lem}

\begin{proof}[Proof of Lemma~\ref{lem:nongaussian_nonindep}]
    \comment{calculations with characteristic functions?}
\end{proof}

Is there a converse? In other words,
does independence of segregation noise imply a Gaussian distribution of trait values?
It is not clear to us in how much generality such a statement might be made,
but we can at least prove the following.
Divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[i]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[i]} 
    \qquad \text{ and } \qquad
    V_M^{[i]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[i]} ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, so
\begin{align*}
    Z_M = U_M^{[1]} + V_M^{[2]} \qquad \text{ and } \qquad
    R_M = \frac{U_M^{[1]} - V_M^{[1]} - U_M^{[2]} + V_M^{[2]}}{ 2 } .
\end{align*}
It turns out that independence of the portions inherited and not inherited by the offspring
implies a Gaussian distribution:

\begin{lem}\label{lem:parental_contribs}
    Suppose that
    $(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ converge in distribution as $M \to \infty$
    to independent random variables $(U^{[1]}, V^{[1]}, U^{[2]}, V^{[2]})$
    (and so $Z_M$, $R_M$, $Z_M^{[1]}$ and $Z_M^{[2]}$ converge to $Z$, $R$, $Z^{[1]}$ and $Z^{[2]}$
    defined in the natural way).
    Then if $R$ is independent of $Z^{[1]}$ and $Z^{[2]}$,
    $(U^{[1]}, V^{[1]}, U^{[2]}, V^{[2]})$ are jointly Gaussian.
\end{lem}

\begin{proof}[Proof of Lemma~\ref{lem:parental_contribs}]
    The Kac-Bernstein theorem states that if two independent random variables
    have the property that if their sum and difference are independent of each other,
    then the random variables have a Gaussian distribution
    (\citet{kac1939characterization,bernstein1941property},
    but see \citet{kagan1973characterization} for many much more general results).
    The assumptions imply that $U^{[1]}$ and $V^{[1]}$
    are such a pair of random variables, and therefore Gaussian.
\end{proof}

It is clear in the Gaussian case
that $U^{[1]}$ and $V^{[1]}$ are asymptotically independent.
More generally, independence depends on how the model is set up.
If the allelic effects $\eta_\ell^{[i]}$ are i.i.d.,
then even for finite $M$, the parental contributions
$(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ are independent.
If the distribution of $\eta_\ell^{[i]}$ is chosen to obtain a non-Gaussian limit,
then Lemma~\ref{lem:parental_contribs} implies that $R$ cannot be independent of the parental traits.

@@

\subsection{Notes}

Let $x_\ell$ and $p_\ell$ be sequences of numbers.
When is, for given $s_1$, $s_2$, $t_1$, and $t_2$,
\begin{align*}
    &
    \prod_{\ell=1}^M
    \left\{
    \frac{1}{2}
        \left( 1 + p_\ell(e^{i s_1 x_\ell/M^{1/\alpha}} - 1) \right)
        \left( 1 + p_\ell(e^{i s_2 x_\ell/M^{1/\alpha}} - 1) \right)
        +
    \frac{1}{2}
        \left( 1 + p_\ell(e^{i t_1 x_\ell/M^{1/\alpha}} - 1) \right)
        \left( 1 + p_\ell(e^{i t_2 x_\ell/M^{1/\alpha}} - 1) \right)
    \right\}
\\ & \qquad \approx
    \prod_{\ell=1}^M
        \left( 1 + p_\ell (e^{i s_1 x_\ell/M^{1/\alpha}} - 1)/2 \right)
        \left( 1 + p_\ell (e^{i s_2 x_\ell/M^{1/\alpha}} - 1)/2 \right)
        \left( 1 + p_\ell (e^{i t_1 x_\ell/M^{1/\alpha}} - 1)/2 \right)
        \left( 1 + p_\ell (e^{i t_2 x_\ell/M^{1/\alpha}} - 1)/2 \right)
\end{align*}

>>>>

Is there a converse? In other words,
does independence of segregation noise imply a Gaussian distribution of trait values?
It is not clear to us in how much generality such a statement might be made,
but we can at least prove the following.
Divide the parental loci into groups according to the results of inheritance,
defining
\begin{align*}
    U_M^{[i]} = \sum_{\ell=1}^M X_\ell \eta_\ell^{[i]} 
    \qquad \text{ and } \qquad
    V_M^{[i]} = \sum_{\ell=1}^M (1-X_\ell) \eta_\ell^{[i]} ,
\end{align*}
so that 
$Z_M^{[1]} = U_M^{[1]} + V_M^{[1]}$ and
$Z_M^{[2]} = U_M^{[2]} + V_M^{[2]}$, so
\begin{align*}
    Z_M = U_M^{[1]} + V_M^{[2]} \qquad \text{ and } \qquad
    R_M = \frac{U_M^{[1]} - V_M^{[1]} - U_M^{[2]} + V_M^{[2]}}{ 2 } .
\end{align*}
It turns out that independence of the portions inherited and not inherited by the offspring
combined with independence of segregation noise and midparent value
implies a Gaussian distribution:

\begin{lem}\label{lem:parental_contribs}
    Suppose that $\{\eta_\ell^{[i]}\}$ are independent for all $\ell$ and $i$,
    that the distribution of $\eta_\ell^{[i]}$ only depends on $\ell$,
    and that $\{X_\ell\}$ are independent of $\{\eta_\ell^{[i]}\}$.
    If the segregation noise is independent of the midparent value
    (i.e., $R_M$ is independent of $(Z^{[1]} + Z^{[2]})/2$)
    then $R_M$ is Gaussian.
\end{lem}

\begin{proof}[Proof of Lemma~\ref{lem:parental_contribs}]
    Note that if inheritance (the $X_\ell$) is independent of the per-locus effects ($\eta_\ell^{[i]}$),
    the latter are all independent of each other,
    and all effects at a given locus are drawn from the same distribution,
    then the inherited allele, 
    $\eta_\ell^{[1]} X_\ell + \eta_\ell^{[2]} (1-X_\ell)$
    is independent of (and has the same distribution as)
    the non-inherited allele,
    $\eta_\ell^{[1]} (1-X_\ell) + \eta_\ell^{[2]} X_\ell$
    -- they are simply two independent draws from the distribution of effects at locus $\ell$.
    Therefore, the sum of inherited alleles, $U_M^{[1]} + V_M^{[2]}$
    are independent of the non-inherited alleles, $U_M^{[2]} + V_M^{[1]}$.

    Now, the Kac-Bernstein theorem states that if two independent random variables
    have the property that if their sum and difference are independent of each other,
    then the random variables have a joint Gaussian distribution
    (\citet{kac1939characterization,bernstein1941property},
    but see \citet{kagan1973characterization} for many much more general results).
    Since $R_M$ is (half) the difference and $(Z^{[1]} + Z^{[2]})/2$ is (half) the sum
    of $U_M^{[1]} + V_M^{[2]}$ and $U_M^{[2]} + V_M^{[1]}$,
    both are therefore jointly Gaussian.
\end{proof}

Since models can be easily set up for which $Z$ is not Gaussian,
the content of this is that for such models, independence of segregation noise
is not likely a good assumption.
Note that the proof does not depend on independence of the $X_\ell$:
indeed, alleles may be inherited together,
although in practice this would quickly create dependencies between nearby allelic effects.
Also note that it is \emph{not} usually true that the components --
for instance, $U_M^{[1]}$ and $U_M^{[2]}$ --
are independent of each other.

\comment{remove the following?}
It is clear in the Gaussian case
that $U^{[1]}$ and $V^{[1]}$ are asymptotically independent.
More generally, independence depends on how the model is set up.
If the allelic effects $\eta_\ell^{[i]}$ are i.i.d.,
then even for finite $M$, the parental contributions
$(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ are independent.
If the distribution of $\eta_\ell^{[i]}$ is chosen to obtain a non-Gaussian limit,
then Lemma~\ref{lem:parental_contribs} implies that $R$ cannot be independent of the parental traits.

@@
